---
title: "ADA Final Project"
author: "Samia Hadji"
date: "2024-11-07"
output: html_document
---


# installing and loading packages
```{r}
pacman::p_load(tidyverse, readr, table1, DiagrammeR, rsvg, ggplot2, tidyverse,haven, dplyr, magrittr, knitr, GGally, survival, ggfortify, survminer, lmtest, foreign, nnet, broom, MASS)
# tidyverse is for data management
# readr is for reading in the csv file
# DiagrammeR is for generating Figure 1
# table1 is for generating table 1
# rsvg for exporting figure 1
```



# 1. Importing Dataset
```{r}
# read dataset imported from SEER
 Project_rawdata <- read.csv("C:/Users/samia/OneDrive/Desktop/ADA Project/SEER ADA Project FINAL FINAL DATASET.txt", header=TRUE)
 
  #colnames(Project_rawdata) <- as.character(Project_rawdata[1, ])
  View(Project_rawdata)
```

# 2.2. Determine: **a.** the number of observations and **b.** the number of variables in the dataframe.Number of Female and Male cases
```{r}
# a and b determine number of observations and variables in raw dataset
dim(Project_rawdata)
# there are 588263 observations and 7 variables
str(Project_rawdata)
table(Project_rawdata$Sex) # 4281 Male cases will be excluded
Project_rawdata$Sex

```

# 3. **a.** Rename the column variables to something shorter and that is descriptive of what the variable is about and **b.** Display that they are renamed.
```{r}
# use names function to rename the 27 variables in order
names(Project_rawdata)[1:7]<-c("Sex", "Age", "Race_Ethnicity", "Combined_Summary_Stage", "Death_Class", "Survival_Months", "Income_range")
# b. display the renamed variables
colnames(Project_rawdata)
```
#4. Determine and display the number of factor, integer, numerical, and character variables there are in the dataset.
```{r}
table(sapply(Project_rawdata, class)) # Using table and sapply functions
summary(Project_rawdata)
```
# 5. Conducting further data management
```{r}

# Recode Survival_Months as numeric

# a. look at values for Survival_Months
Project_rawdata$Survival_Months

# b.convert to numeric for statistics
Project_rawdata$Survival_Months<-as.numeric(Project_rawdata$Survival_Months)
Project_rawdata$Survival_Months # check values 


Project_rawdata


```

6. Recode race Ethnicity as factor variables, replace the ranking with White as reference group(0), recode Non-Hispanic Unknown Race as Missing

```{r}
table(Project_rawdata$Race_Ethnicity)
Project_rawdata <- Project_rawdata %>%
  mutate(Race_Ethnicity = case_when(Race_Ethnicity == "Non-Hispanic White" ~ 0,
                          Race_Ethnicity == "Non-Hispanic Black" ~ 1,
                          Race_Ethnicity == "Non-Hispanic American Indian/Alaska Native" ~ 2,
                          Race_Ethnicity == "Non-Hispanic Asian or Pacific Islander" ~ 3,
                          Race_Ethnicity == "Hispanic (All Races)" ~ 4,
                          Race_Ethnicity == "Non-Hispanic Unknown Race" ~ NA))

Project_rawdata$Race_Ethnicity <- factor(Project_rawdata$Race_Ethnicity, levels = c(0:4), labels = c("Non-Hispanic White", "Non-Hispanic Black", "Non-Hispanic American Indian/Alaska Native", "Non-Hispanic Asian or Pacific Islander", "Hispanic (All Races)"))
table(Project_rawdata$Race_Ethnicity,useNA = "always")


```


```{r}

# create a new variable using Income recoded into categories, create new variable using Age recoded into fewer groups,Create a new variable using Death_Classification recoded into 0 and 1(for the purpose of Survival analysis that will be conducted), and recode missing/uknown/unstaged into NA

Project_data <- Project_rawdata %>%
  mutate(Income_category = ifelse(`Income_range` == "< $40,000", "Lower income", 
                                  ifelse(`Income_range` %in% c("$90,000 - $94,999", "$95,000 - $99,999", "$100,000 - $109,999", "$110,000 - $119,999", "$120,000+"), "Upper income", 
                                         ifelse(`Income_range` == "Unknown/missing/no match/Not 1990-2022", NA, "Middle income")))) %>%
  mutate(Age_group = ifelse(Age %in% c("15-19 years", "01-04 years", "10-14 years", "05-09 years", "00 years"), "<20",
                            ifelse(Age %in% c("25-29 years", "20-24 years"), "20-29",
                                   ifelse(Age %in% c("35-39 years", "30-34 years"), "30-39",
                                          ifelse(Age %in% c("45-49 years", "40-44 years"), "40-49",
                                                 ifelse(Age %in% c("55-59 years", "50-54 years"), "50-59",
                                                        ifelse(Age %in% c("60-64 years", "65-69 years"), "60-69",
                                                               ifelse(Age %in% c("75-79 years", "70-74 years"), "70-79", ">80")))))))) %>%
  mutate(Death_Classified = ifelse(Death_Class == "Alive or dead of other cause", 0,
                              ifelse(Death_Class == "Dead (attributable to this cancer dx)", 1, NA)))



# replace missing/unknown/unstaged values with NAs
Project_data <- Project_data %>%
 mutate(Combined_Summary_Stage = ifelse(Combined_Summary_Stage %in% c("Unknown/unstaged", "In situ"), NA, Combined_Summary_Stage)) %>%
  mutate(Death_Class = ifelse(Death_Class %in% c("Dead (missing/unknown COD)"), NA, Death_Class)) %>%
  mutate(Death_Classified = ifelse(Death_Classified %in% c("Dead (missing/unknown COD)"), NA, Death_Classified))



# Check variables for correct categorization 
table(Project_data$'Income_range', Project_data$Income_category, useNA = "always")
table(Project_data$Death_Class, Project_data$Death_Classified, useNA = "always")
table(Project_data$Age, Project_data$Age_group, useNA = "always")
table(Project_data$Combined_Summary_Stage, useNA = "always")

```



# Figure 1a; drop na one variable at the time, flow chart
- Let's next make a figure 1 that shows exclusions to arrive at your analytic data set. 
- We can do this with a package called DiagrammeR and the function grViz.


# First we will make the following exclusions one at a time
- Exclusion 1: exclude Male cases
```{r}
Project_data1 <- Project_data %>%
  filter(Sex == "Female") 

ex1 <- nrow(Project_data)- nrow(Project_data1)
ex1
```
- Exclusion 2 those missing data on Race_Ethnicity
```{r}
table(Project_data2$Race_Ethnicity, useNA = "always")

Project_data2 <- Project_data1 %>%
  drop_na(Race_Ethnicity)



ex2 <- nrow(Project_data1) - nrow(Project_data2)
ex2
```


- Exclusion 3 those with missing data on Income_category
```{r}
Project_data3 <- Project_data2 %>%
  drop_na(Income_category) 

ex3 <- nrow(Project_data2) - nrow(Project_data3)
ex3
```
- Exclusion 4 those with missing data on Death_Classified


```{r}
Project_data4 <- Project_data3 %>%
  drop_na(Death_Classified) 

ex4 <- nrow(Project_data3) - nrow(Project_data4) 
ex4
```



- Exclusion 5 those with missing data on Combined_Summary_Stage
```{r}
Project_data5 <- Project_data4 %>%
  drop_na(Combined_Summary_Stage) 

ex5 <- nrow(Project_data4) - nrow(Project_data5) 
ex5
```


- Exclusion 6 those with missing data on Survival_Months

```{r}
Project_data6 <- Project_data5 %>%
  drop_na(Survival_Months) 

ex6 <- nrow(Project_data5) - nrow(Project_data6) 
ex6

```
# Exclusion 7 # let's make another dataframe for the prpose of conducting Survival analysis that excludes Male cases, with missing data on Income_Category, with missing data on Death Classified, and with missing data on survival time (survmo) only, without excluding missing on Combined Stage Summary. Figure 2b with branching will be generated

```{r}
Project_data7 <- Project_data4 %>%
  drop_na(Survival_Months) 

ex7 <- nrow(Project_data4) - nrow(Project_data7) 
ex7
```
# Let's use the grViz function from the DiagrammeR package to make a flow chart of how we arrived at our analytic dataframe

```{r}
figure1a <- grViz(diagram = "digraph flowchart{ # gives beginning of flowchart

      # node definitions with substituted label text
      
      node [fontname = Helvetica, shape = rectangle, fontsize=10] 
      
      # Define the nodes: nodes indicate how many boxes you will have in your diagram. Since I have three sets of exclusions, I will have four nodes.
      
      node1 [label = '@@1'] # starting number
      node2 [label = '@@2'] # number after exclusion 1
      node3 [label = '@@3'] # number after exclusion 2
      node4 [label = '@@4'] # number after exclusion 3
      node5 [label = '@@5'] # number after exclusion 4
      node6 [label = '@@6'] # number after exclusion 5
      node7 [label = '@@7'] # number after exclusion 6
      
      
      # edge definitions with the node IDs are used to indicate how the rectangle boxes flow from each other. 
      
      node1 -> node2 -> node3 -> node4 -> node5 -> node6 -> node7
      
      
}
      # This set of code provides the text in each rectangle box.
      [1]: 'Records received from SEER for Breast cancer diagnoses n = 588263'
      [2]: 'Excluding 4281 Male 2007 n = 583,982'
      [3]: 'Excluding 4283 individuals with missing data on \\n Race Ehnicity n = 579,699'
      [4]: 'Excluding 59 individuals with missing data on \\n Income category n = 579,640'
      [5]: 'Excluding 1517 individuals with missing data on \\n Death classification n = 578,123'
      [6]: 'Excluding 11537 individuals with missing data on \\n Combined Summary Stage n = 566,586'
      [7]: 'Excluding 52 individuals with missing data on \\n Survival Months n = 566,534'
      ")
figure1a

```

# Suppose we also conduct an analysis on survival where we don't need to exclude those with missing Combined Summary Stage at diagnosis: 
- we can use branching to do this
```{r}
# note in this function all text is green but you can still comment out text with hashtags
# for more information on grViz that uses the "DOT" language, see: https://rich-iannone.github.io/DiagrammeR/articles/graphviz-mermaid.html

figure1b <- grViz(diagram = "digraph flowchart{ # gives beginning of flowchart

      # node definitions with substituted label text
      
      node [fontname = Helvetica, shape = rectangle, fontsize=10] 
      
      # Define the nodes: nodes indicate how many boxes you will have in your diagram. Since I have three sets of exclusions, I will have four nodes.
      
      node1 [label = '@@1'] # starting number
      node2 [label = '@@2'] # number after exclusion 1
      node3 [label = '@@3'] # number after exclusion 2
      node4 [label = '@@4'] # number after exclusion 3
      node5 [label = '@@5'] # number after exclusion 4
      node6 [label = '@@6'] # number after exclusion 5
      node7 [label = '@@7'] # number after exclusion 6
      
      # edge definitions with the node IDs are used to indicate how the rectangle boxes flow from each other. 
      
      node1 -> node2 -> node3 -> node4 -> node5
      node5 -> node6 # branching
      node5 -> node7 # branching
      
}

      # This set of code provides the text in each rectangle box.
      [1]: 'Records received from SEER for Breast cancer diagnoses n = 588263'
      [2]: 'Excluding 4281 Male 2007 n = 583,982'
      [3]: 'Excluding 4283 individuals with missing data on Race Ethnicity n = 579,699'
      [4]: 'Excluding 59 individuals with missing data on \\n Income Category n = 579,640'
      [5]: 'Excluding 1517 individuals with missing data on \\n Death Classification n = 578,123'
      [6]: 'Excluding 11537 individuals with missing data on \\n Combined Summary Stage n = 566,586'
      [7]: 'Excluding 2706 individuals with missing data on \\n Survival Months n = 575,417'
      ")
figure1b
```
# export figure

```{r}
figure1a %>%
  DiagrammeRsvg::export_svg() %>% 
  charToRaw() %>% 
  rsvg::rsvg_pdf("Figure 1.pdf")
```

# Table 1

```{r}
#Drop missing on Death class.........
Project_data_clean <- Project_data6 %>%
  drop_na(Death_Class)

# Adding labels to the variables because the raw variables are not publication-quality. This is important in producing a high-quality table.
label(Project_data_clean$Age_group)<-"Age group"
label(Project_data_clean$Race_Ethnicity)<-"Race Ethnicity"
label(Project_data_clean$Income_category)<-"Income Category"
label(Project_data_clean$Combined_Summary_Stage)<-"Combined Stage Summary"
label(Project_data_clean$Death_Class)<-"Death Classification"
label(Project_data_clean$Survival_Months)<-"Survival time (months)"




# Generate Table 1
tbl1 <- table1(~ Age_group + Race_Ethnicity + Income_category + Combined_Summary_Stage + Death_Class + Survival_Months,overall="Total", footnote='SD = standard deviation', rowlabelhead="Variable", data = Project_data_clean)
tbl1
```
# Stratified table 1 by Race Ethnicity
```{r}
#Generate Table1 stratified results by race ethnicity status

table1(~ Age_group + Income_category + Combined_Summary_Stage  + Death_Class + Survival_Months|Race_Ethnicity, overall="Total", footnote='SD = standard deviation', 
       rowlabelhead="Variable", Project_data_clean)
```



# Figure 2

```{r}

ggp <- Project_data_clean %>%
  ggplot(aes(x = Race_Ethnicity, y = after_stat(prop), fill = Combined_Summary_Stage, by = Race_Ethnicity)) +
  geom_bar(position = position_dodge(), stat = "prop") +
  geom_text(aes(label = paste0(round(100 * after_stat(prop), 1), "%")),
    position = position_dodge(.9), stat = "prop", vjust = -.2
  ) +
  labs(x = "Race Ethnicity Groups", y = "Proportion", title = "Combined Summary Stage by Race Ethnicity (n = 657; 
       SEER, 2011-2021)",
       fill = "Combined Summary Stage")
ggp + scale_x_discrete(guide=guide_axis(angle=45),labels = function(x) str_wrap(x, width = 10))




```



#Kaplan Meier


```{r}
#overall
# overall
overall <- survfit(Surv(Survival_Months, Death_Classified) ~ 1, Project_data6) 
# get survival probabilities
summary(overall)
 
# Race
KM_Race <- survfit(Surv(Survival_Months, Death_Classified ) ~ Race_Ethnicity, Project_data6)
summary(KM_Race)
# Income
KM_Income <- survfit(Surv(Survival_Months, Death_Classified) ~ Income_category, Project_data6)
summary(KM_Income)
```

# Survival probabilities

```{r}
#ggsurvplot Overall
summary(overall)
ggsurvplot(overall, Project_data6 = Project_data6, conf.int=TRUE, pval = TRUE,  tables.theme = clean_theme())
#ggsurvplot Race Ethnicity
ggsurvplot(KM_Race, Project_data6, Conf.int= TRUE, pval = TRUE,
             legend = c(0.5, 0.5),  # Adjust legend position (optional, in x/y proportions)
  legend.title = "Race Ethnicity") # Customize legend title
#ggsurvplot Income Category
ggsurvplot(KM_Income, Project_data6, Conf.int= TRUE, pval = TRUE,
             legend = c(0.5, 0.5),  # Adjust legend position (optional, in x/y proportions)
  legend.title = "Income_category") # Customize legend title

  
```

# 9. Get median survival time overall, by sex, and by race (note if the results are unexpected, think about why)

```{r}
overall
KM_Race
KM_Income
```

# 10. Calculate exact log rank in R for the difference in survival curves between Race Ethnicity subroups, Income categories

```{r}
# Race
survdiff(Surv(Survival_Months, Death_Classified) ~ Race_Ethnicity, data=Project_data6)

# Income
survdiff(Surv(Survival_Months, Death_Classified) ~ Income_category, data=Project_data6)
```



```{r}
#Conduct pairwise differences test for race that has more than two categories.
pairwise_survdiff(Surv(Survival_Months, Death_Classified) ~ Race_Ethnicity, data=Project_data6) # post-hoc test
```

```{r}
#Conduct pairwise differences test for Income that has more than two categories.
pairwise_survdiff(Surv(Survival_Months, Death_Classified) ~ Income_category, data=Project_data6) # post-hoc test
```


## adjust for Income_category
## Stratified log rank test
- The observed minus expected scores are calculated within strata of each group and then summed across strata (see excel sheet 4 for intuition).
- The stratified log rank test allows for testing of whether survival curves differ controlling for the stratified variable
- Let's control for Income category using the strata function

```{r}
survdiff(Surv(Survival_Months, Death_Classified) ~ Race_Ethnicity + strata(Income_category), data=Project_data6) # adjusted log-rank
```
# Cox propor

```{r}
## Which group has a higher hazard of Death by Breast cancer (within race and ethnicity groups)? Run univariate Cox model to examine the association between race and ethnicity group and death by breast cancer. Interpret the results.

race.model <- coxph(Surv(Survival_Months, Death_Classified)~Race_Ethnicity, Project_data6, ties="efron") 
summary(race.model)

# Interpretation: Those who are of such race ethnicity  had a ..... (95% CI ....-.....) times higher hazard of death by breast cancer than those of .....race ethnicity.

```

# Adjust the Cox model for Income, combined Summary Stage, and age 

```{r}
race_adj.model<-coxph(Surv(Survival_Months, Death_Classified) ~ Race_Ethnicity + Income_category + Age_group + Combined_Summary_Stage , Project_data6)
summary(race_adj.model)
```

 ## Compare nested models using the likelihood ratio test (a measure of model fit) and interpret the findings
 
```{r}
lrtest(race.model, race_adj.model)
```
 #Include an interaction term in the model between Race Ethnicity and Income categories to see if there is effect modification of the hazard of dying of breast cancer in those among the different race and ethnicity groups according to their Income category
 
```{r}
race_int.model1<-coxph(Surv(Survival_Months, Death_Classified)~Race_Ethnicity  + Income_category + Income_category*Race_Ethnicity, Project_data6)
summary(race_int.model1)
library(broom)
# Set scipen to a high value to avoid scientific notation
options(scipen = 999)
data <- as.data.frame(tidy(race_int.model1, exponentiate = TRUE, conf.int = TRUE))
```
Call:
coxph(formula = Surv(Survival_Months, Death_Classified) ~ Race_Ethnicity + 
    Income_category + Income_category * Race_Ethnicity, data = Project_data6)

  n= 566534, number of events= 51302 

                                                                                           coef
Race_EthnicityNon-Hispanic Black                                                       0.352242
Race_EthnicityNon-Hispanic American Indian/Alaska Native                              -0.181430
Race_EthnicityNon-Hispanic Asian or Pacific Islander                                  -0.031904
Race_EthnicityHispanic (All Races)                                                     0.116287
Income_categoryMiddle income                                                          -0.358654
Income_categoryUpper income                                                           -0.574080
Race_EthnicityNon-Hispanic Black:Income_categoryMiddle income                          0.199013
Race_EthnicityNon-Hispanic American Indian/Alaska Native:Income_categoryMiddle income  0.248328
Race_EthnicityNon-Hispanic Asian or Pacific Islander:Income_categoryMiddle income     -0.117063
Race_EthnicityHispanic (All Races):Income_categoryMiddle income                        0.005573
Race_EthnicityNon-Hispanic Black:Income_categoryUpper income                           0.226837
Race_EthnicityNon-Hispanic American Indian/Alaska Native:Income_categoryUpper income   0.507322
Race_EthnicityNon-Hispanic Asian or Pacific Islander:Income_categoryUpper income      -0.156287
Race_EthnicityHispanic (All Races):Income_categoryUpper income                        -0.050659
                                                                                      exp(coef)
Race_EthnicityNon-Hispanic Black                                                       1.422253
Race_EthnicityNon-Hispanic American Indian/Alaska Native                               0.834077
Race_EthnicityNon-Hispanic Asian or Pacific Islander                                   0.968600
Race_EthnicityHispanic (All Races)                                                     1.123318
Income_categoryMiddle income                                                           0.698616
Income_categoryUpper income                                                            0.563223
Race_EthnicityNon-Hispanic Black:Income_categoryMiddle income                          1.220198
Race_EthnicityNon-Hispanic American Indian/Alaska Native:Income_categoryMiddle income  1.281881
Race_EthnicityNon-Hispanic Asian or Pacific Islander:Income_categoryMiddle income      0.889529
Race_EthnicityHispanic (All Races):Income_categoryMiddle income                        1.005589
Race_EthnicityNon-Hispanic Black:Income_categoryUpper income                           1.254625
Race_EthnicityNon-Hispanic American Indian/Alaska Native:Income_categoryUpper income   1.660837
Race_EthnicityNon-Hispanic Asian or Pacific Islander:Income_categoryUpper income       0.855313
Race_EthnicityHispanic (All Races):Income_categoryUpper income                         0.950602
                                                                                       se(coef)
Race_EthnicityNon-Hispanic Black                                                       0.071797
Race_EthnicityNon-Hispanic American Indian/Alaska Native                               0.238615
Race_EthnicityNon-Hispanic Asian or Pacific Islander                                   0.578546
Race_EthnicityHispanic (All Races)                                                     0.143569
Income_categoryMiddle income                                                           0.037745
Income_categoryUpper income                                                            0.038793
Race_EthnicityNon-Hispanic Black:Income_categoryMiddle income                          0.073046
Race_EthnicityNon-Hispanic American Indian/Alaska Native:Income_categoryMiddle income  0.247246
Race_EthnicityNon-Hispanic Asian or Pacific Islander:Income_categoryMiddle income      0.579029
Race_EthnicityHispanic (All Races):Income_categoryMiddle income                        0.144386
Race_EthnicityNon-Hispanic Black:Income_categoryUpper income                           0.078196
Race_EthnicityNon-Hispanic American Indian/Alaska Native:Income_categoryUpper income   0.265319
Race_EthnicityNon-Hispanic Asian or Pacific Islander:Income_categoryUpper income       0.579128
Race_EthnicityHispanic (All Races):Income_categoryUpper income                         0.146485
                                                                                            z
Race_EthnicityNon-Hispanic Black                                                        4.906
Race_EthnicityNon-Hispanic American Indian/Alaska Native                               -0.760
Race_EthnicityNon-Hispanic Asian or Pacific Islander                                   -0.055
Race_EthnicityHispanic (All Races)                                                      0.810
Income_categoryMiddle income                                                           -9.502
Income_categoryUpper income                                                           -14.798
Race_EthnicityNon-Hispanic Black:Income_categoryMiddle income                           2.724
Race_EthnicityNon-Hispanic American Indian/Alaska Native:Income_categoryMiddle income   1.004
Race_EthnicityNon-Hispanic Asian or Pacific Islander:Income_categoryMiddle income      -0.202
Race_EthnicityHispanic (All Races):Income_categoryMiddle income                         0.039
Race_EthnicityNon-Hispanic Black:Income_categoryUpper income                            2.901
Race_EthnicityNon-Hispanic American Indian/Alaska Native:Income_categoryUpper income    1.912
Race_EthnicityNon-Hispanic Asian or Pacific Islander:Income_categoryUpper income       -0.270
Race_EthnicityHispanic (All Races):Income_categoryUpper income                         -0.346
                                                                                                  Pr(>|z|)
Race_EthnicityNon-Hispanic Black                                                               0.000000929
Race_EthnicityNon-Hispanic American Indian/Alaska Native                                           0.44705
Race_EthnicityNon-Hispanic Asian or Pacific Islander                                               0.95602
Race_EthnicityHispanic (All Races)                                                                 0.41795
Income_categoryMiddle income                                                          < 0.0000000000000002
Income_categoryUpper income                                                           < 0.0000000000000002
Race_EthnicityNon-Hispanic Black:Income_categoryMiddle income                                      0.00644
Race_EthnicityNon-Hispanic American Indian/Alaska Native:Income_categoryMiddle income              0.31520
Race_EthnicityNon-Hispanic Asian or Pacific Islander:Income_categoryMiddle income                  0.83978
Race_EthnicityHispanic (All Races):Income_categoryMiddle income                                    0.96921
Race_EthnicityNon-Hispanic Black:Income_categoryUpper income                                       0.00372
Race_EthnicityNon-Hispanic American Indian/Alaska Native:Income_categoryUpper income               0.05586
Race_EthnicityNon-Hispanic Asian or Pacific Islander:Income_categoryUpper income                   0.78726
Race_EthnicityHispanic (All Races):Income_categoryUpper income                                     0.72947
                                                                                         
Race_EthnicityNon-Hispanic Black                                                      ***
Race_EthnicityNon-Hispanic American Indian/Alaska Native                                 
Race_EthnicityNon-Hispanic Asian or Pacific Islander                                     
Race_EthnicityHispanic (All Races)                                                       
Income_categoryMiddle income                                                          ***
Income_categoryUpper income                                                           ***
Race_EthnicityNon-Hispanic Black:Income_categoryMiddle income                         ** 
Race_EthnicityNon-Hispanic American Indian/Alaska Native:Income_categoryMiddle income    
Race_EthnicityNon-Hispanic Asian or Pacific Islander:Income_categoryMiddle income        
Race_EthnicityHispanic (All Races):Income_categoryMiddle income                          
Race_EthnicityNon-Hispanic Black:Income_categoryUpper income                          ** 
Race_EthnicityNon-Hispanic American Indian/Alaska Native:Income_categoryUpper income  .  
Race_EthnicityNon-Hispanic Asian or Pacific Islander:Income_categoryUpper income         
Race_EthnicityHispanic (All Races):Income_categoryUpper income                           
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

                                                                                      exp(coef)
Race_EthnicityNon-Hispanic Black                                                         1.4223
Race_EthnicityNon-Hispanic American Indian/Alaska Native                                 0.8341
Race_EthnicityNon-Hispanic Asian or Pacific Islander                                     0.9686
Race_EthnicityHispanic (All Races)                                                       1.1233
Income_categoryMiddle income                                                             0.6986
Income_categoryUpper income                                                              0.5632
Race_EthnicityNon-Hispanic Black:Income_categoryMiddle income                            1.2202
Race_EthnicityNon-Hispanic American Indian/Alaska Native:Income_categoryMiddle income    1.2819
Race_EthnicityNon-Hispanic Asian or Pacific Islander:Income_categoryMiddle income        0.8895
Race_EthnicityHispanic (All Races):Income_categoryMiddle income                          1.0056
Race_EthnicityNon-Hispanic Black:Income_categoryUpper income                             1.2546
Race_EthnicityNon-Hispanic American Indian/Alaska Native:Income_categoryUpper income     1.6608
Race_EthnicityNon-Hispanic Asian or Pacific Islander:Income_categoryUpper income         0.8553
Race_EthnicityHispanic (All Races):Income_categoryUpper income                           0.9506
                                                                                      exp(-coef)
Race_EthnicityNon-Hispanic Black                                                          0.7031
Race_EthnicityNon-Hispanic American Indian/Alaska Native                                  1.1989
Race_EthnicityNon-Hispanic Asian or Pacific Islander                                      1.0324
Race_EthnicityHispanic (All Races)                                                        0.8902
Income_categoryMiddle income                                                              1.4314
Income_categoryUpper income                                                               1.7755
Race_EthnicityNon-Hispanic Black:Income_categoryMiddle income                             0.8195
Race_EthnicityNon-Hispanic American Indian/Alaska Native:Income_categoryMiddle income     0.7801
Race_EthnicityNon-Hispanic Asian or Pacific Islander:Income_categoryMiddle income         1.1242
Race_EthnicityHispanic (All Races):Income_categoryMiddle income                           0.9944
Race_EthnicityNon-Hispanic Black:Income_categoryUpper income                              0.7971
Race_EthnicityNon-Hispanic American Indian/Alaska Native:Income_categoryUpper income      0.6021
Race_EthnicityNon-Hispanic Asian or Pacific Islander:Income_categoryUpper income          1.1692
Race_EthnicityHispanic (All Races):Income_categoryUpper income                            1.0520
                                                                                      lower .95
Race_EthnicityNon-Hispanic Black                                                         1.2356
Race_EthnicityNon-Hispanic American Indian/Alaska Native                                 0.5225
Race_EthnicityNon-Hispanic Asian or Pacific Islander                                     0.3117
Race_EthnicityHispanic (All Races)                                                       0.8478
Income_categoryMiddle income                                                             0.6488
Income_categoryUpper income                                                              0.5220
Race_EthnicityNon-Hispanic Black:Income_categoryMiddle income                            1.0574
Race_EthnicityNon-Hispanic American Indian/Alaska Native:Income_categoryMiddle income    0.7896
Race_EthnicityNon-Hispanic Asian or Pacific Islander:Income_categoryMiddle income        0.2859
Race_EthnicityHispanic (All Races):Income_categoryMiddle income                          0.7577
Race_EthnicityNon-Hispanic Black:Income_categoryUpper income                             1.0764
Race_EthnicityNon-Hispanic American Indian/Alaska Native:Income_categoryUpper income     0.9874
Race_EthnicityNon-Hispanic Asian or Pacific Islander:Income_categoryUpper income         0.2749
Race_EthnicityHispanic (All Races):Income_categoryUpper income                           0.7134
                                                                                      upper .95
Race_EthnicityNon-Hispanic Black                                                         1.6372
Race_EthnicityNon-Hispanic American Indian/Alaska Native                                 1.3314
Race_EthnicityNon-Hispanic Asian or Pacific Islander                                     3.0103
Race_EthnicityHispanic (All Races)                                                       1.4884
Income_categoryMiddle income                                                             0.7523
Income_categoryUpper income                                                              0.6077
Race_EthnicityNon-Hispanic Black:Income_categoryMiddle income                            1.4080
Race_EthnicityNon-Hispanic American Indian/Alaska Native:Income_categoryMiddle income    2.0812
Race_EthnicityNon-Hispanic Asian or Pacific Islander:Income_categoryMiddle income        2.7671
Race_EthnicityHispanic (All Races):Income_categoryMiddle income                          1.3345
Race_EthnicityNon-Hispanic Black:Income_categoryUpper income                             1.4624
Race_EthnicityNon-Hispanic American Indian/Alaska Native:Income_categoryUpper income     2.7936
Race_EthnicityNon-Hispanic Asian or Pacific Islander:Income_categoryUpper income         2.6612
Race_EthnicityHispanic (All Races):Income_categoryUpper income                           1.2667

Concordance= 0.564  (se = 0.001 )
Likelihood ratio test= 3074  on 14 df,   p=<0.0000000000000002
Wald test            = 3302  on 14 df,   p=<0.0000000000000002
Score (logrank) test = 3412  on 14 df,   p=<0.0000000000000002

 
```{r}
lrtest(race_adj.model, race_int.model1)
```
 
 # check for influential observations: assess whether any observations are exerting influence on the coefficient estimates.
 
 
```{r}
ggcoxdiagnostics(race_int.model1, type = "dfbeta", sline=FALSE, ggtheme = theme_bw()) 
# the pattern should not change the beta by a large degree. Here the betas change by a factor of less than +/-0.2 for all covariates
```
 
 

 
 #Testing the PH assumption
 ## 1. Log-log curves: log(-log(survival)) against log(time)
 
```{r}
Project_data.surv <- survfit(Surv(Project_data6$Survival_Months, Project_data6$Death_Classified) ~ Race_Ethnicity, Project_data6) # get survival probabilities
summary(Project_data.surv)

plot(Project_data.surv , col=c("black", "red"), main= "rx log log plot", fun="cloglog", xlab="Time", ylab="log(-log(S(t))") ##ph assumption not violated........
```

# 2. Schoenfeld goodness-of-fit test 


```{r}
test.ph<-cox.zph(race_adj.model, terms=FALSE) 
test.ph
```

# 3. Schoenfeld residuals plots


```{r}

plot(test.ph, resid=FALSE)

#  Here is yet another way to do the Schoenfeld residual plots by time, showing the p-values from the Schoenfeld Test for each covariate
ggcoxzph(test.ph)
```


# Running Multinominal Logistic Regression: Race Ehnicity vs Combined Stage Summary
## recode variables for analysis

```{r}
# Combined_Stage_Summary - change it to numeric, set Localized as reference
Project_data6 <- Project_data6 %>%
  mutate(Combined_Summary_Stage_new = case_when(Combined_Summary_Stage=="Localized" ~ 0,
                              Combined_Summary_Stage=="Regional" ~ 1,
                              Combined_Summary_Stage=="Distant" ~ 2)) 
Project_data6$Combined_Summary_Stage_new <- factor(Project_data6$Combined_Summary_Stage_new,
                                                   levels = c(0:2), labels = c("Localized", "Regional", "Distant")) # make it a factor and give factor labels for different  combined summary stage subcategories

head(Project_data6)
```

## running the multinominal regression

```{r}
# running the model
model_multi <- multinom(Project_data6$Combined_Summary_Stage_new ~ Project_data6$Race_Ethnicity, data=Project_data6)
summary(model_multi)
results <- tidy(model_multi, exponentiate = TRUE, conf.int = TRUE)
write.csv(results, "C:/Users/samia/OneDrive/Desktop/ADA Project/multinom_results.csv")
```
 

.

##  Just for exploring;Adjusting for age: The reason is: the effect of Age on the detection and management of Breast cance.

```{r}
model_Adj <- multinom(Combined_Summary_Stage_new ~ Race_Ethnicity + Age_group, data=Project_data6)
summary(model_Adj)

## extract coefficients from the model, exponentiate and CI
tidy(model_Adj, conf.int=TRUE, exponentiate = TRUE)
```


















